\documentclass[11pt,twoside]{article}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsthm}
\usepackage{amsfonts}
\usepackage{microtype}
\usepackage{hyperref}
\usepackage{graphicx}
\usepackage{tikz}
\usepackage{mathrsfs}

\newtheorem{theorem}{Theorem}
\newtheorem{lemma}{Lemma}
\newtheorem{proposition}{Proposition}
\newtheorem{definition}{Definition}
\newtheorem{conjecture}{Conjecture}
\newtheorem{remark}{Remark}

\title{\textbf{Spectral Methods in Data Compression and Geometric Encoding: \\
Block Entropy, Information Conservation, and Generalized Surfaces of Revolution}}
\author{\textbf{Padraic} \\ \textit{Independent Researcher}}
\date{\today}

\begin{document}

\maketitle

\begin{abstract}
We present three interconnected investigations at the intersection of information theory, signal processing, and differential geometry. First, we formalize the observation that grouping a binary string into blocks of size $n$ and computing the resulting $n$-gram entropy reveals a spectral structure: the block size that minimizes the per-bit entropy corresponds to the dominant period of the string, recoverable in $\mathcal{O}(N \log N)$ time via the Fast Fourier Transform. Second, we provide a rigorous analysis of why encoding binary data onto continuous curves---whether sinusoidal trajectories or high-degree polynomials---and applying Fourier series compression necessarily fails, tracing the obstruction to a fundamental conservation of degrees of freedom and the floating-point inflation penalty. Third, we develop the theory of generalized surfaces of revolution, wherein a target curve is rotated about a curved axis parameterized by the Frenet--Serret frame expressed through unit quaternions in $\mathbb{H}$, establishing a structural connection to the $\operatorname{SU}(2)$ sandwich operator of quantum mechanics. We show that taking the winding number to infinity yields a spiral space-filling path that reduces the dimensionality of the surface from $\mathbb{R}^3$ to a single $1$-dimensional signal amenable to classical spectral compression.
\end{abstract}

%=============================================================================
\section{Introduction}
%=============================================================================

The central thread unifying the three investigations in this paper is the interplay between \emph{discrete information} and \emph{continuous representation}. Shannon's source coding theorem establishes a hard lower bound on the expected codeword length for lossless compression. Yet the bound applies to a specific probability model, and by changing the model---grouping symbols into blocks, shifting alignment, or transforming the representation space---one can appear to violate the bound while in fact operating under a different (and tighter) entropy measure.

This observation motivates Part~I, where we formalize the block entropy framework and show that the FFT provides an efficient mechanism for finding the natural ``fundamental frequency'' of a binary string. Part~II examines the converse question: can we \emph{gain} compressibility by embedding discrete data into a continuous domain? We show rigorously that the answer is no, due to an information-theoretic conservation law. Part~III departs from compression to develop a geometric framework---generalized surfaces of revolution around curved axes---that unexpectedly reconnects to compression through a spiral dimensional reduction that \emph{does} produce a compressible continuous signal from a 3D surface.

%=============================================================================
\part*{I. Block Entropy and FFT-Based Compression}
%=============================================================================

\section{First-Order Shannon Entropy and Its Limitations}

Let $\mathbf{b} = (b_1, b_2, \dots, b_N)$ be a binary string of length $N$, with $b_i \in \{0, 1\}$. The first-order Shannon entropy of $\mathbf{b}$ is
\begin{equation}
	H_1(\mathbf{b}) = -p_0 \log_2 p_0 - p_1 \log_2 p_1,
\end{equation}
where $p_0$ and $p_1$ are the empirical frequencies of $0$ and $1$ in $\mathbf{b}$. This quantity assumes the source is \emph{memoryless}: the probability of each bit is independent of all others.

If $\mathbf{b}$ possesses any sequential structure---repeating patterns, Markov dependencies, or periodic motifs---$H_1$ completely fails to capture it. A string such as $\mathbf{b} = 01010101\dots$ has $p_0 = p_1 = 1/2$ and thus $H_1 = 1$ bit, yet it is trivially compressible because the entire string is determined by a single bit and a period.

\section{Block Entropy and Symbol Extension}

\begin{definition}[Block Entropy]
	For a positive integer $n$ dividing $N$, partition $\mathbf{b}$ into $N/n$ consecutive non-overlapping blocks of $n$ bits. Each block is interpreted as a symbol from the alphabet $\Sigma_n = \{0, 1\}^n$ of cardinality $2^n$. Let $p_\sigma^{(n)}$ denote the empirical frequency of symbol $\sigma \in \Sigma_n$. The \emph{$n$-gram block entropy} is
	\begin{equation}
		H_n(\mathbf{b}) = -\sum_{\sigma \in \Sigma_n} p_\sigma^{(n)} \log_2 p_\sigma^{(n)}.
	\end{equation}
	The \emph{per-bit block entropy} is $h_n(\mathbf{b}) = H_n(\mathbf{b}) / n$.
\end{definition}

\begin{proposition}[Monotonicity of Per-Bit Block Entropy]
	For any stationary source, the per-bit block entropy is non-increasing:
	\begin{equation}
		h_n(\mathbf{b}) \leq h_{n-1}(\mathbf{b}) \leq \cdots \leq h_1(\mathbf{b}) = H_1(\mathbf{b}).
	\end{equation}
	Equality holds throughout if and only if the source is memoryless.
\end{proposition}

\begin{remark}
	By sweeping $n$ over candidate block sizes and evaluating $h_n$, one searches for the \emph{natural word length} of the data---the block size at which the per-bit entropy is minimized. If the string has an underlying period of $T$ bits, then $n = T$ (or a multiple thereof) will yield a maximally skewed symbol distribution and thus the lowest $h_n$.
\end{remark}

\subsection{Cyclic Shifts and Padding}

When $n \nmid N$, we extend $\mathbf{b}$ by cyclic repetition (looping the string on itself) and pad with zeros to the nearest multiple of $n$. Let $s \in \{0, 1, \dots, n-1\}$ denote a bit-level shift applied before blocking. The per-bit entropy becomes a function of both block size and shift:
\begin{equation}
	h_{n,s}(\mathbf{b}) = \frac{1}{n} H\bigl(\text{BlockPartition}(\text{Shift}_s(\mathbf{b}), n)\bigr).
\end{equation}
The optimal compression parameters are
\begin{equation}
	(n^*, s^*) = \operatorname*{arg\,min}_{n, s} \; h_{n,s}(\mathbf{b}).
\end{equation}

\subsection{Metadata Overhead}

To decompress the string, the decoder requires: the chosen block size $n^*$, the shift $s^*$, and the Huffman dictionary for the $n^*$-bit alphabet. For short strings (say $N < 100$), this metadata can dominate the total compressed size. However, the overhead is $\mathcal{O}(\log N + 2^{n^*})$ bits, which is amortized as $N \to \infty$.

\section{The FFT as a Period Detector}

Rather than brute-force search over all $(n, s)$ pairs, we exploit the duality between periodicity detection and spectral analysis.

\subsection{Signal Conversion}

Map the binary string to a zero-mean real signal:
\begin{equation}
	x_i = 2b_i - 1, \qquad x_i \in \{-1, +1\}.
\end{equation}

\subsection{Autocorrelation via the Wiener--Khinchin Theorem}

The autocorrelation $R(\tau) = \sum_{i} x_i \, x_{i+\tau}$ measures how well the signal matches a time-shifted copy of itself. By the Wiener--Khinchin theorem, $R(\tau)$ is the inverse Fourier transform of the power spectral density:
\begin{equation}
	R(\tau) = \mathcal{F}^{-1}\!\left[\, |\hat{X}(f)|^2 \,\right](\tau),
\end{equation}
where $\hat{X}(f) = \mathcal{F}[x](f)$ is the discrete Fourier transform.

\begin{theorem}[FFT Period Detection]
	Let $\hat{X}(f)$ be the DFT of the signal $(x_1, \dots, x_N)$, and let $f_{\max}$ be the frequency bin maximizing the power spectral density $|\hat{X}(f)|^2$ (excluding the DC component $f = 0$). Then the dominant period of the string is
	\begin{equation}
		T = \frac{N}{f_{\max}},
	\end{equation}
	and the optimal block size is $n^* = \operatorname{round}(T)$.
\end{theorem}

\subsection{Phase Extraction for Optimal Shift}

The FFT yields complex coefficients $\hat{X}(f) = A(f) e^{i\phi(f)}$. The phase angle $\phi(f_{\max})$ at the dominant frequency encodes the alignment of the periodic pattern relative to the start of the string:
\begin{equation}
	s^* = \operatorname{round}\!\left(\frac{\phi(f_{\max})}{2\pi} \cdot T\right) \bmod T.
\end{equation}

This extracts both optimal parameters in $\mathcal{O}(N \log N)$ time, compared to $\mathcal{O}(N \cdot T_{\max}^2)$ for brute-force search.

\subsection{The Goertzel Optimization}

When the candidate block sizes are restricted to powers of $2$ (i.e., $n \in \{2, 4, 8, 16, \dots\}$), one need only evaluate the power spectrum at $\mathcal{O}(\log N)$ specific frequency bins. The Goertzel algorithm computes the DFT at a single target frequency in $\mathcal{O}(N)$ time, yielding a total complexity of $\mathcal{O}(N \log N)$ for the restricted search.

%=============================================================================
\part*{II. Information Conservation Under Domain Transforms}
%=============================================================================

\section{Motivation: Encoding Binary Data on Continuous Curves}

A natural question arises: if the Fourier transform can compress audio signals and images by discarding negligible high-frequency coefficients, can we embed a binary string into a continuous function and compress it via Fourier series?

We examine two specific encoding schemes and prove that both necessarily \emph{expand} the data rather than compress it.

\section{Scheme A: Slope-Modulated Sinusoidal Encoding}

\subsection{Construction}

Given $N$ bits, partition them into $4$ chunks of $N/4$ bits, one per quadrant of a sinusoidal template. Consider the first quadrant: we construct a piecewise-linear path from $(0, 0)$ to $(\pi/2, 1)$ using $N/4$ discrete steps. At each step $i$:
\begin{itemize}
	\item If $b_i = b_{i-1}$ (no transition), the slope remains unchanged.
	\item If $b_i \neq b_{i-1}$ (transition), the slope decreases by $\frac{4}{N}$.
\end{itemize}
Starting from maximal slope at the origin, the boundary conditions are enforced so the curve passes through $(\pi/2, 1)$, $(\pi, 0)$, $(3\pi/2, -1)$, and $(2\pi, 0)$---matching $\sin(x)$ at multiples of $\pi/2$.

\begin{remark}
	A uniform string (all identical bits) produces a straight line from $(0,0)$ to $(\pi/2, 1)$. A maximally alternating string produces a piecewise-linear approximation of $\sin(x)$.
\end{remark}

\subsection{Why Fourier Compression Fails}

The resulting piecewise-linear curve has $N/4$ ``knuckle points'' where the slope changes. Each knuckle is a discontinuity in the first derivative.

\begin{theorem}[Gibbs Obstruction]
	To reconstruct a piecewise-linear curve with $K$ slope discontinuities via Fourier series to sufficient precision for lossless bit recovery, at least $K$ non-negligible Fourier coefficients are required. Each coefficient is a real number requiring at least 32 bits of floating-point storage.
\end{theorem}

\begin{proof}[Proof sketch]
	The curve has $K = N/4$ degrees of freedom (the $K$ slope choices). The Fourier transform is a unitary (information-preserving) map. Therefore, the frequency-domain representation has at least $K$ significant coefficients. Storing $K$ floating-point numbers at 32 bits each yields $32K = 8N$ bits---a factor of $8\times$ expansion over the original $N$ bits.
\end{proof}

\section{Scheme B: Polynomial Interpolation Followed by Fourier Analysis}

\subsection{Construction}

Instead of connecting the slope-determined points with line segments, fit a smooth polynomial through all $K = N/4$ data points. This eliminates derivative discontinuities.

\subsection{Runge's Phenomenon and Degree Explosion}

A polynomial passing through $K$ arbitrary points has degree at most $K - 1$, requiring $K$ coefficients $(a_0, a_1, \dots, a_{K-1})$. The polynomial oscillates violently between data points (Runge's phenomenon), generating massive high-frequency energy in the Fourier domain.

\begin{theorem}[Information Conservation Law]
	Let $\mathbf{b} \in \{0,1\}^N$ be a binary string and let $\Phi: \{0,1\}^N \to \mathbb{R}^M$ be any injective (lossless) map into a space of real-valued parameters. Then:
	\begin{equation}
		M \geq \frac{N}{\lceil \log_2(1/\epsilon) \rceil},
	\end{equation}
	where $\epsilon$ is the minimum precision required per real parameter to distinguish all $2^N$ possible inputs. For standard IEEE-754 floating-point representations, the storage cost is $32M \geq 32N / \lceil \log_2(1/\epsilon) \rceil$ bits.
\end{theorem}

In practice, the precision $\epsilon$ required to distinguish $2^N$ inputs is exponentially small, so $M \geq N$ and the total storage is at least $32N$ bits.

\begin{remark}[Connection to Telecommunications]
	While Fourier-based encoding fails as a \emph{compressor}, it is precisely the mathematical framework used for modern wireless data \emph{transmission}. OFDM (Orthogonal Frequency-Division Multiplexing) and QAM (Quadrature Amplitude Modulation) map binary bits onto the amplitudes and phases of sinusoidal subcarriers---not to compress data, but to transmit it efficiently over analog channels.
\end{remark}

%=============================================================================
\part*{III. Generalized Surfaces of Revolution via Quaternionic Frames}
%=============================================================================

\section{Classical Surfaces of Revolution}

A classical surface of revolution is generated by rotating a \emph{target curve} $C(s)$ about a straight-line axis. If $C(s) = (f(s), 0, s)$ and the axis is the $z$-axis, the resulting surface is parameterized by
\begin{equation}
	\mathbf{r}(s, \theta) = \bigl(f(s)\cos\theta, \; f(s)\sin\theta, \; s\bigr), \qquad \theta \in [0, 2\pi).
\end{equation}

We seek to generalize this construction to allow rotation about a \emph{curved} axis.

\section{The Curved Axis and Moving Frame}

\begin{definition}[Axis Curve]
	Let $\gamma: [a, b] \to \mathbb{R}^3$ be a smooth, regular curve parameterized by arc length $s$, with curvature $\kappa(s) > 0$ and torsion $\tau(s)$.
\end{definition}

At each point $\gamma(s)$, we define the Frenet--Serret frame $\{\mathbf{T}(s), \mathbf{N}(s), \mathbf{B}(s)\}$, satisfying
\begin{equation}
	\frac{d}{ds}
	\begin{pmatrix} \mathbf{T} \\ \mathbf{N} \\ \mathbf{B} \end{pmatrix}
	=
	\begin{pmatrix} 0 & \kappa & 0 \\ -\kappa & 0 & \tau \\ 0 & -\tau & 0 \end{pmatrix}
	\begin{pmatrix} \mathbf{T} \\ \mathbf{N} \\ \mathbf{B} \end{pmatrix}.
\end{equation}

\begin{definition}[Target Curve and Radius Function]
	Let $C$ be a second curve. At each $s \in [a, b]$, the normal line from $\gamma(s)$ along $\mathbf{N}(s)$ intersects $C$ at a point whose distance from $\gamma(s)$ defines the \emph{radius function} $r(s)$.
\end{definition}

\begin{definition}[Non-Degeneracy Condition]
	The construction is non-degenerate when the maximum radius does not exceed the minimum radius of curvature of the axis:
	\begin{equation}
		\max_{s \in [a,b]} r(s) \leq \min_{s \in [a,b]} \frac{1}{\kappa(s)}.
	\end{equation}
	This prevents the normal lines from intersecting each other, which would cause the coordinate system to fold.
\end{definition}

\section{Generalized Surface of Revolution}

The generalized surface is parameterized by
\begin{equation}
	\mathbf{r}(s, \theta) = \gamma(s) + r(s)\bigl[\cos\theta \; \mathbf{N}(s) + \sin\theta \; \mathbf{B}(s)\bigr], \qquad \theta \in [0, 2\pi).
\end{equation}

When $\gamma$ is a straight line ($\kappa = 0$), $\mathbf{N}$ and $\mathbf{B}$ are constant and this reduces to the classical surface of revolution. For a curved axis, the Frenet frame rotates along the path, ``twisting'' the surface according to the torsion $\tau(s)$.

\section{Quaternionic Parameterization}

\subsection{Unit Quaternions and SU(2)}

Rather than tracking three vectors $(\mathbf{T}, \mathbf{N}, \mathbf{B})$ independently, we encode the entire Frenet frame in a single unit quaternion $q(s) \in \mathbb{H}$, $\|q\| = 1$.

\begin{definition}[Quaternionic Frame Equation]
	The Frenet--Serret equations are equivalent to the quaternionic differential equation
	\begin{equation}
		\frac{dq}{ds} = \frac{1}{2} q(s) \, \omega(s),
	\end{equation}
	where $\omega(s) = \tau(s)\,\mathbf{i} + \kappa(s)\,\mathbf{k}$ is the pure quaternion encoding the curvature and torsion.
\end{definition}

\subsection{The Sandwich Operator and Quantum Mechanics}

To rotate a 3D vector $\mathbf{v}$ (embedded as the pure quaternion $v = 0 + v_x\mathbf{i} + v_y\mathbf{j} + v_z\mathbf{k}$), we apply:
\begin{equation}
	v' = q \, v \, q^{-1}.
\end{equation}

\begin{remark}[Connection to Quantum Observables]
	This is structurally identical to the quantum mechanical expectation value $\langle \hat{A} \rangle = \langle \psi | \hat{A} | \psi \rangle$. Unit quaternions map to $\operatorname{SU}(2)$ (the double cover of $\operatorname{SO}(3)$), so tracing a path $q(s)$ along the axis curve is equivalent to tracking the continuous unitary evolution of a spin-$\frac{1}{2}$ state. The target curve behaves as an observable whose expectation value evolves along the axis.
\end{remark}

\section{The Inverse Problem: Unwrapping to a Cylinder}

\subsection{Finding the Straightening Axis}

Given a target curve $C_1(s)$, we ask: \emph{what axis curve $\gamma(s)$ produces a constant radius $r(s) = c$?}

If $r(s) = c$ is constant, then $C_1(s) = \gamma(s) + c \, \mathbf{N}(s)$ is an \emph{offset curve} (or parallel curve) of $\gamma$. The required axis is the \emph{evolute} of $C_1$---the locus of centers of curvature:
\begin{equation}
	\gamma(s) = C_1(s) - \frac{1}{\kappa_{C_1}(s)} \, \mathbf{N}_{C_1}(s).
\end{equation}

\subsection{Application: 3D Model Compression}

Consider a 3D model (e.g., a human head) represented in cylindrical coordinates $(r, \theta, z)$. Slicing in the $zr$-plane at each angle $\theta$ produces a set of 2D curves $\{C_\theta\}$. For each such slice, solving the inverse problem yields an axis curve $\gamma_\theta$ that ``unwraps'' the complex organic geometry into a cylinder. Chaining this operation across all slices transforms the full 3D surface into a collection of cylinders and smooth residual surfaces.

\begin{remark}[Connection to Offset Curve Convergence]
	This operation is related to a classical result in the geometry of curves: given a convex closed curve, the iterated offset curve (drawn at constant distance in every direction) converges to a circle. Similarly, iterating the ``unwrap to cylinder'' operation successively smooths any convex surface.
\end{remark}

\section{The Spiral Limit and Dimensional Reduction}

\subsection{From Discrete Wraps to Continuous Surfaces}

Consider a single helical path wrapping around the generalized surface with $n$ periods between the endpoints $a$ and $b$:
\begin{equation}
	\mathbf{c}_n(t) = \gamma(s(t)) + r(s(t))\bigl[\cos(n \cdot 2\pi t) \; \mathbf{N}(s(t)) + \sin(n \cdot 2\pi t) \; \mathbf{B}(s(t))\bigr],
\end{equation}
where $t \in [0, 1]$ and $s(t) = a + (b - a)t$.

At $n = 1$, this is a single loop. As $n$ increases, the spiral wraps tighter and tighter. In the limit $n \to \infty$, the image of $\mathbf{c}_n$ is dense in the surface, analogous to a space-filling curve constrained to a 2D manifold.

\begin{proposition}[Dimensional Reduction]
	The spiral path $\mathbf{c}_n(t)$ reduces the surface to a $1$-dimensional signal: any point on the surface is identified by a single parameter $t \in [0, 1]$, representing arc-length along the spiral.
\end{proposition}

\subsection{Compression via the Spiral Transform}

Sampling the radius $r$ along the spiral path produces a $1$-dimensional waveform $r(t)$.

\begin{itemize}
	\item A perfect cylinder yields $r(t) = \text{const}$: a single Fourier coefficient.
	\item A smooth vase yields a slowly varying $r(t)$: a few low-frequency coefficients suffice.
	\item A detailed organic surface (face, head) yields $r(t)$ with high-frequency ripples at feature locations (nose, eyes), but with strong spatial correlation throughout.
\end{itemize}

Because organic 3D surfaces have immense spatial correlation (the nose blends smoothly into the cheek), the Fourier transform of $r(t)$ concentrates energy in the low-frequency coefficients. High-frequency coefficients decay rapidly and can be discarded with negligible perceptual loss.

\begin{remark}
	This resolves the apparent contradiction with Part~II. Fourier compression fails for \emph{discrete, independent} binary data because the degrees of freedom do not decay spectrally. But for \emph{continuous, correlated} geometric data, the Fourier transform exploits spatial correlation to achieve genuine compression. The spiral transform converts a 3D surface into exactly the type of smooth, correlated signal that Fourier analysis was designed to compress.
\end{remark}

\subsection{Connection to Fourier Analysis on Lie Groups}

Because quaternions are non-commutative ($\mathbf{i}\mathbf{j} \neq \mathbf{j}\mathbf{i}$), the limit of infinitesimal rotations along the axis does not yield a standard Fourier decomposition. Instead, it maps onto the \emph{Spherical Fourier Transform} or Fourier analysis on the Lie group $\operatorname{SU}(2)$.

In this framework, the 3D shape is decomposed not into flat sines and cosines, but into \emph{Wigner $D$-matrices}---the irreducible representations of $\operatorname{SU}(2)$:
\begin{equation}
	f(\alpha, \beta, \gamma) = \sum_{\ell=0}^{\infty} \sum_{m=-\ell}^{\ell} \sum_{m'=-\ell}^{\ell} \hat{f}^{\ell}_{mm'} \, D^{\ell}_{mm'}(\alpha, \beta, \gamma),
\end{equation}
where the Euler angles $(\alpha, \beta, \gamma)$ parameterize the rotation. These are the same spherical harmonics used in quantum mechanics to describe electron orbital angular momentum.

%=============================================================================
\section{Conclusion and Unifying Perspective}
%=============================================================================

The three parts of this paper trace a common arc: the search for the right \emph{representation space} in which structure becomes compressible.

In Part~I, we showed that a binary string's hidden periodicity is invisible to first-order entropy but revealed by block entropy, and that the FFT efficiently identifies the optimal block size---the ``fundamental frequency'' of the data. In Part~II, we proved that embedding discrete data into continuous domains cannot circumvent the conservation of degrees of freedom: information is neither created nor destroyed by a change of basis, and the floating-point overhead inflates the representation. In Part~III, we demonstrated that for data which \emph{already} lives in a continuous, correlated domain (3D geometry), a carefully constructed spiral dimensional reduction---parameterized by quaternionic frames that obey the same algebra as quantum spin---produces exactly the type of smooth, spectrally concentrated signal that Fourier analysis excels at compressing.

The unifying principle is that \emph{compression is possible when and only when the representation captures the correlations of the source}. Block coding captures sequential patterns. Fourier analysis captures smooth spatial correlations. The quaternionic spiral transform captures the rotational symmetry of surfaces. Each tool succeeds precisely in the domain for which it was designed, and fails when applied across that boundary.

\end{document}
